---
title: "Wildfire Data"
author: "Laina Cleaton"
date: "2025-10-26"
output: html_document
---

```{r setup_tidyverse, eval = TRUE}
library(tidyverse)
```

```{r read_files, eval = TRUE}
evac_zone_status <- read_csv("evac_zone_status_geo_event_map.csv")
evac_zones <- read_csv("evac_zones_gis_evaczone.csv")
geo_events <- read_csv("geo_events_geoevent.csv")
```

```{r evac_zone_status, eval = FALSE}
view(evac_zone_status)
```

```{r evac_zones, eval = FALSE}
view(evac_zones)
```

```{r geo_events, eval = FALSE}
view(geo_events)
```

In order to conduct a temporal and spatial analysis, we will join the **evac_zone_status**, **geo_events**, and **evac_zones** datasets. The 'changelog' datasets are unnecessary for the analysis.

The first join, **evac_zone_status** and **geo_events** (by geo_event_id), is important because **evac_zone_status** does not include information on the wildfire (name, geographical location, etc.). **evac_zone_status** only tracks the evacuation zone's activity over time, so by joining these two tables, we will be able to see the evacuation status changes associated with a specific wildfire.

The second join, **evac_zone_status** and **evac_zones** (by uid_v2), brings in the spatial component. Each evacuation zone's geometry and display name will be attached to its status updates, which means we'll be able to pinpoint the when and where each evacuation occurred. Evacuation boundaries can be visualized, their areas calculated, and they can be overlayed with demographic data.

Altogether, the "core spatial dataset" brings together the components of the who, when, and where of wildfire evacuations; which fire the evacuations were associated with, when they were created, and where they occurred.

```{r install_sf, eval = FALSE}
install.packages("sf")
```

```{r wildfire_joins, eval = TRUE}

library(sf)

# --- Convert CSVs to sf objects using WKT column ---
# evac_zones
evac_zones <- st_as_sf(evac_zones, wkt = "geom", crs = 4326)

# select evac_zone_status columns
evac_zone_status_sel <- evac_zone_status %>%
  rename(status_date = date_created) %>%
  select(uid_v2, geo_event_id, status_date)

# select geo_events columns
geo_events_sel <- geo_events %>%
  rename(geo_event_id = id) %>%
  select(geo_event_id, name, lat, lng)

# select evac_zones columns
evac_zones_sel <- evac_zones %>%
  rename(geom_evac_zone = geom) %>%
  select(uid_v2, display_name, geom_evac_zone)

#join columns
core_wildfire_dataset <- evac_zone_status_sel %>%
  left_join(geo_events_sel, by = "geo_event_id") %>%
  left_join(evac_zones_sel, by = "uid_v2") %>%
  st_as_sf()

```

We cannot use rows with NA in **display_name** because they lack geometry, and could break spatial joins. Therefore, we can filter these NA's out.

```{r filter_core_wildfire_dataset, eval = TRUE}
core_wildfire_geo <- core_wildfire_dataset %>%
  filter(!is.na(display_name))
```

After joining the selected WatchDuty tables into a single dataset, each record now represents a single evacuation status update. This means that there can be multiple entries per evacuation zone, **uid_v2**, as its status changes over time. Thus we will need to summarize those updates into meaningful evacuation timelines:

```         
1. Evacuation start time: the earliest status_date recorded for that zone. 
2. Evacuation end time: the latest status_date recorded.
3. Evacuation duration: the time difference between those two points.
```

This step outputs one row per evacuation zone per fire. Thus we can describe how long communities were under evacuation and visualize the temporal progression of the evacuations.

```{r evac_timeline, eval = TRUE}
evac_timeline <- core_wildfire_geo %>%
  st_drop_geometry() %>%      
  group_by(uid_v2) %>%
  summarise(
    n_updates = n(),
    min_date = min(status_date),
    max_date = max(status_date),
    diff_hours = as.numeric(difftime(max_date, min_date, units = "hours"))
  ) %>%
  arrange(desc(diff_hours))

```


Now I need to check the validity of my geom polygons, to ensure I can filter the data and visualize the results.


```{r polygon_validity_check, eval = TRUE}
library(sf)

# Convert the CSV with WKT geometry to sf safely
evac_zones_sf <- st_as_sf(evac_zones, wkt = "geom", crs = 4326, quiet = TRUE)

# Rename geometry column for clarity
evac_zones_sf <- evac_zones_sf %>%
  rename(geometry = geom)

# Fix invalid polygons 
evac_zones_sf <- st_make_valid(evac_zones_sf)

# Drop empty geometries
evac_zones_sf <- evac_zones_sf %>%
  filter(!st_is_empty(geometry))

# Check
print(st_is_valid(evac_zones_sf) %>% table())

```

I have 37324 valid geometries, and 134 invalid geometries, even after conversion. Now I can filter out those invalid geometries.

```{r geometry_filter, eval = TRUE}
evac_zones_sf <- evac_zones_sf %>%
  filter(st_is_valid(geometry))
```

```{r joins_fixed, eval = TRUE}
evac_zone_status_sel <- evac_zone_status %>%
  rename(status_date = date_created) %>%
  select(uid_v2, geo_event_id, status_date)

geo_events_sel <- geo_events %>%
  rename(geo_event_id = id) %>%
  select(geo_event_id, name, lat, lng)

evac_zones_sel <- evac_zones_sf %>%
  select(uid_v2, display_name, geometry)

core_wildfire_geo <- evac_zone_status_sel %>%
  left_join(geo_events_sel, by = "geo_event_id") %>%
  left_join(evac_zones_sel, by = "uid_v2") %>%
  st_as_sf()
```


Next, I wanted to see the distribution of the proportion of all the WatchDuty wildfires. I wanted to see which duration times made up the majority of these evacuations. 

```{r evac_timeline_filter_summary, eval = TRUE}
evac_timeline <- evac_timeline %>%
  mutate(diff_hours = as.numeric(diff_hours)) %>%
  filter(!is.na(diff_hours), diff_hours > 0)
```


```{r evac_timeline_distribution, eval = TRUE}
library(tidyverse)
ggplot(evac_timeline, aes(x = diff_hours)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    bins = 30,
    fill = "steelblue",
    color = "white",
    alpha = 0.7) +
  labs(
    title = "Evacuation Duration Distribution",
    subtitle = "Proportion of all Evacuation Zones",
    x = "Evacuation Duration (hours)",
    y = "Proportion of Evacuation Zones"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal()
```

[explanation of summary statistics]
As we can see, most of the wildfires were under 3,000 hours. I wanted to see how this overall distribution compares to just wildfires in California.


```{r california_filter, eval = TRUE}
library(sf)

ca_bbox <- st_sfc(
  st_polygon(list(rbind(
    c(-125, 32),
    c(-125, 42),
    c(-114, 42),
    c(-114, 32),
    c(-125, 32)
  ))),
  crs = st_crs(core_wildfire_geo)
)

cali_wildfire_geo <- core_wildfire_geo[st_intersects(core_wildfire_geo, ca_bbox, sparse = FALSE), ]
```

```{r cali_evac_timeline_distribution, eval = TRUE}
library(scales)

evac_timeline <- evac_timeline %>%
  mutate(diff_hours = as.numeric(diff_hours)) %>%
  filter(!is.na(diff_hours), diff_hours > 0)

cali_evac_timeline <- cali_wildfire_geo %>%
  st_drop_geometry() %>%
  group_by(uid_v2) %>%
  summarise(
    n_updates = n(),
    min_date = min(status_date),
    max_date = max(status_date),
    diff_hours = as.numeric(difftime(max_date, min_date, units = "hours"))
  ) %>%
  filter(!is.na(diff_hours), diff_hours > 0)

cali_evac_timeline %>%
  summarise(
    n_zones = n(),
    mean_duration = mean(diff_hours, na.rm = TRUE),
    median_duration = median(diff_hours, na.rm = TRUE),
    sd_duration = sd(diff_hours, na.rm = TRUE),
    min_duration = min(diff_hours, na.rm = TRUE),
    max_duration = max(diff_hours, na.rm = TRUE),
    p25 = quantile(diff_hours, 0.25, na.rm = TRUE),
    p75 = quantile(diff_hours, 0.75, na.rm = TRUE)
  )

ggplot(cali_evac_timeline, aes(x = diff_hours)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),   # convert counts to proportions
    bins = 30,
    fill = "darkorange",
    color = "white",
    alpha = 0.7
  ) +
  labs(
    title = "Evacuation Duration Distribution",
    subtitle = "California Evacuation Zones",
    x = "Evacuation Duration (hours)",
    y = "Percentage of California Evacuation Zones"
  ) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal(base_size = 13)
```

[explanation of summary statistics]
This distribution is much different. We can see that the California evacuation zones have a much higher evacuation duration on average than the distribution of all locations. We can also see that there are 222 evacuation zones located in California, which is about 49.55% of the total 448 evacuation zones recorded. 

```{r cali_vs_all_distribution, eval = TRUE}

library(scales)

evac_timeline <- evac_timeline %>%
  mutate(diff_hours = as.numeric(diff_hours)) %>%
  filter(!is.na(diff_hours), diff_hours > 0)

evac_timeline_full <- evac_timeline %>%
  mutate(region = "All Zones") %>%
  bind_rows(
    cali_evac_timeline %>%
      mutate(region = "California")
  )

ggplot(evac_timeline_full, aes(x = diff_hours, fill = region)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    bins = 30,
    alpha = 0.5,   # transparency so both histograms are visible
    position = "identity",  # overlay instead of stacked
    color = "white"
  ) +
  scale_fill_manual(values = c("All Zones" = "steelblue", "California" = "darkorange")) +
  labs(
    title = "Evacuation Duration Distribution",
    subtitile = "California vs All Zones",
    x = "Evacuation Duration (hours)",
    y = "Percentage of Evacuation Zones",
    fill = "Region"
  ) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal(base_size = 13)
```



Next, we can visualize the joined dataset, using leaflet. Although California is my place of interest, I was curious about where other fires were located. 


```{r install_leaflet, eval = FALSE}
install.packages("leaflet")
```


```{r leaflet_joins_visualization, eval = FALSE}
library(leaflet)

leaflet(filter(core_wildfire_geo, name %in% top_fires)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~factor(name),
    color = "#444",
    weight = 1,
    popup = ~paste("<b>Fire:</b>", name, "<br><b>Zone:</b>", display_name)
  )
```

This shows that the fires from WatchDuty data were located in Minnesota (near Duluth, on the western tip of Lake Superior) and California. 


Next I wanted to observe the distribution of evacuation durations as a heatmap. 

```{r attach_diff_hours, eval = TRUE} 
core_wildfire_geo_time <- core_wildfire_geo %>%
  left_join(evac_timeline %>% select(uid_v2, diff_hours), by = "uid_v2") %>%
  filter(!is.na(diff_hours))
```

```{r nationwide_heatmap_duration, eval = FALSE}
library(leaflet)

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = core_wildfire_geo_time$diff_hours
)

leaflet(core_wildfire_geo_time) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(diff_hours),
    fillOpacity = 0.7,
    color = "#444",
    weight = 0.3,
    popup = ~paste0(
      "<b>Fire:</b> ", name,
      "<br><b>Zone:</b> ", display_name,
      "<br><b>Duration (hours):</b> ", round(diff_hours, 1)
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~diff_hours,
    title = "Evacuation Duration (Hours)",
    opacity = 0.7
  )
```


```{r add_time_cali, eval = TRUE}
cali_wildfire_geo_time <- cali_wildfire_geo %>%
  left_join(cali_evac_timeline %>% select(uid_v2, diff_hours), by = "uid_v2") %>%
  filter(!is.na(diff_hours))
```

```{r cali_heatmap_duration, eval = FALSE}
pal_ca <- colorNumeric(
  palette = "YlOrRd",
  domain = cali_wildfire_geo_time$diff_hours
)

leaflet(cali_wildfire_geo_time) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_ca(diff_hours),
    fillOpacity = 0.7,
    color = "#444",
    weight = 0.5,
    popup = ~paste0(
      "<b>Fire:</b> ", name,
      "<br><b>Zone:</b> ", display_name,
      "<br><b>Duration (hours):</b> ", round(diff_hours, 1)
    )
  ) %>%
  addLegend(
    pal = pal_ca,
    values = ~diff_hours,
    title = "CA Evacuation Duration (Hours)",
    opacity = 0.7
  )
```

Then I wanted to see which zones in California had the longest evacuation durations. (The top 10%)

```{r longest_evacuation_cali, eval = TRUE}
cali_wildfire_geo_time_clean <- cali_wildfire_geo_time %>%
  filter(!is.na(name), !is.na(display_name))  

# 90th percentile threshold for durations
longest_threshold <- quantile(cali_wildfire_geo_time_clean$diff_hours, 0.90, na.rm = TRUE)


# Filter only the longest evacuations (top 10%) by zone

cali_longest_zones <- cali_wildfire_geo_time_clean %>%
  group_by(uid_v2, display_name) %>%     
  summarise(
    fire_name = first(name),             # fire name for reference
    duration_hours = max(diff_hours)     
  ) %>%
  ungroup() %>%
  filter(duration_hours >= longest_threshold) %>%
  arrange(desc(duration_hours))


# Table of top longest zones

cali_longest_table <- cali_longest_zones %>%
  select(
    Fire = fire_name,
    Zone = display_name,
    Duration_Hours = duration_hours
  )

# View top 15 zones
head(cali_longest_table, 15)

# Map the longest zones in California

library(leaflet)

# Join back with the geometry from your original CA dataset
cali_longest_zones_sf <- cali_wildfire_geo_time_clean %>%
  filter(uid_v2 %in% cali_longest_zones$uid_v2) %>%
  group_by(uid_v2, display_name, name) %>%
  summarise(duration_hours = max(diff_hours)) %>%
  st_as_sf()

```

```{r longest_cali_heatmap, eval=FALSE}
# Create heatmap-style leaflet map
pal_ca <- colorNumeric(
  palette = "Blues",
  domain = cali_longest_zones_sf$duration_hours
)

leaflet(cali_longest_zones_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_ca(duration_hours),
    fillOpacity = 0.8,
    color = "#222",
    weight = 1,
    popup = ~paste0(
      "<b>Fire:</b> ", name,
      "<br><b>Zone:</b> ", display_name,
      "<br><b>Duration (hours):</b> ", round(duration_hours, 1)
    )
  ) %>%
  addLegend(
    pal = pal_ca,
    values = ~duration_hours,
    title = "Evacuation Duration (Hours)",
    opacity = 0.7
  )
```

The analysis focused on the top 10% of evacuation zones by duration in California. These zones represent communities most disrupted by wildfires, regardless of the specific fire event. By examining evacuation durations and spatial extent, we can approximate the impact on residents, workers, and small businesses.


```{r tiger_read, eval = TRUE}
# Load California block groups
ca_bg <- st_read("tl_2025_06_bg.shp")

# Quick check
ca_bg
plot(st_geometry(ca_bg))  # shows all polygons
```

How the TIGER/Line shapefile helps: TIGER/Line gives you polygons for every block group. Each polygon is a small area with a GEOID that can be matched to population or economic data from ACS. (a grid over California that tells you where people live.)

Now we can overlay census data to know roughly how many people were potentially affected in each of the top evacuation zones in California. 

```{r acs_population_data, eval = TRUE}
library(tidycensus)

pop_ca_bg <- get_acs(
  geography = "block group",
  variables = "B01003_001",  # total population
  state = "CA",
  year = 2021,
  survey = "acs5",
  geometry = FALSE
)

# pop_ca_bg now has GEOID + total population (estimate)
head(pop_ca_bg)
```

```{r acs_tiger_join, eval = TRUE}
library(tidyverse)

# Make sure GEOID columns are character
ca_bg$GEOID <- as.character(ca_bg$GEOID)
pop_ca_bg$GEOID <- as.character(pop_ca_bg$GEOID)

# Join population
ca_bg_pop <- ca_bg %>%
  left_join(pop_ca_bg %>% select(GEOID, estimate), by = "GEOID")

# Check
head(ca_bg_pop)
```


```{r evac_zone_populations, eval = TRUE}
library(tidyverse)

# Assume these are your objects:
# ca_bg_pop -> block groups with population
# cali_longest_zones_sf -> evacuation zones polygons

# Make sure both have the same CRS
st_crs(cali_longest_zones_sf) <- st_crs(ca_bg_pop)

# Intersect block groups with evacuation zones
evac_pop_overlay <- st_intersection(
  cali_longest_zones_sf, 
  ca_bg_pop
)

# Estimate population affected per zone
evac_population_summary <- evac_pop_overlay %>%
  group_by(display_name) %>%  # evacuation zone
  summarise(
    affected_population = sum(estimate, na.rm = TRUE)
  ) %>%
  arrange(desc(affected_population))

# Check results
evac_population_summary
```

Now we can visualize it on a map using color to show higher/lower population affected, show community impact, and combine with evacuation duration to estimate the "Cumulative Population Exposure" metric as a measure of community burden.

```{r population_heatmap, eval = FALSE}
library(leaflet)

pal <- colorNumeric("YlOrRd", domain = evac_population_summary$affected_population)

leaflet(evac_population_summary) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(affected_population),
    fillOpacity = 0.7,
    color = "#444",
    weight = 0.5,
    popup = ~paste0(display_name, "<br>Population: ", affected_population)
  ) %>%
  addLegend(
    pal = pal,
    values = ~affected_population,
    title = "Population Affected",
    opacity = 0.7
  )
```

```{r population_duration, eval = TRUE}
library(tidyverse)

# Join the population summary with evacuation duration
evac_impact <- evac_population_summary %>%
  left_join(
    cali_longest_zones_sf %>% 
      st_drop_geometry() %>% 
      select(display_name, duration_hours),
    by = "display_name"
  ) %>%
  mutate(
    cumulative_population_exposure = affected_population * duration_hours
  ) %>%
  arrange(desc(cumulative_population_exposure))

# View top impacted zones
head(evac_impact)
```


```{r population_duration_heatmap, eval = FALSE}

library(leaflet)

# Define color palette based on evacuation duration
pal_duration <- colorNumeric("YlOrRd", domain = evac_impact$duration_hours)

leaflet(evac_impact) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_duration(duration_hours),  # color by duration
    fillOpacity = 0.7,
    color = "#444",
    weight = 0.5,
    popup = ~paste0(
      "<b>Zone:</b> ", display_name, "<br>",
      "<b>Duration (hrs):</b> ", round(duration_hours, 1), "<br>",
      "<b>Population:</b> ", affected_population
    )
  ) %>%
  addLegend(
    pal = pal_duration,
    values = ~duration_hours,
    title = "Evacuation Duration (hrs)",
    opacity = 0.7
  )

```

This heatmap-style map where zones with high numbers of people and long durations stand out shows the community impact. 

```{r population_duration_table, eval = FALSE}
library(tidyverse)

# Create summary table sorted by population affected
evac_summary_table <- evac_impact %>%
  st_drop_geometry() %>%  # remove geometry for table display
  select(
    Zone = display_name,
    Duration_Hours = duration_hours,
    Population_Affected = affected_population
  ) %>%
  arrange(desc(Population_Affected))  # sort by population descending

# View the table
evac_summary_table
```


```{r population_duration_top10_barchart, eval = TRUE}
library(ggplot2)
library(tidyverse)
library(scales)

# Summarize by zone first to remove duplicates
evac_summary <- evac_impact %>%
  group_by(display_name, duration_hours) %>%
  summarise(
    affected_population = sum(affected_population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(affected_population) %>%
  mutate(
    Zone_Label = factor(display_name, levels = display_name)
  )

# Plot
ggplot(evac_summary, aes(x = affected_population, y = Zone_Label, fill = duration_hours)) +
  geom_col() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Duration (hrs)") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Evacuation Zones by Population Affected",
    subtitle = "In California",
    x = "Population Affected",
    y = "Evacuation Zone"
  ) +
  theme_minimal(base_size = 13)

```

How much evacuation time does the average person in this zone experience? 

To answer that question, we need to calculate the Evacuation Duration per person. I decided to report per-capita evacuation duration for all California zones rather than restricting to a percentile cutoff, so the analysis reflects the full distribution of community impacts and avoids arbitrary thresholds.

```{r evacuation_per_person_cali, eval = TRUE}

library(tidyverse)

california_evac_summary <- evac_impact %>%
  st_drop_geometry() %>%
  mutate(
    Evacuation_Duration_Per_Person = duration_hours / affected_population
  ) %>%
  select(
    Zone = display_name,
    Duration_Hours = duration_hours,
    Population_Affected = affected_population,
    Evacuation_Duration_Per_Person
  ) %>%
  arrange(desc(Evacuation_Duration_Per_Person))

# View table
california_evac_summary
```

```{r duration_per_person_cali_barchart, eval = TRUE}
library(ggplot2)
library(tidyverse)
library(scales)

# Summarize by zone 
evac_summary <- evac_impact %>%
  group_by(display_name, duration_hours) %>%
  summarise(
    duration_hours = max(duration_hours, na.rm = TRUE), affected_population = sum(affected_population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Evacuation_Duration_Per_Person = duration_hours / affected_population,
    Zone_Label = factor(display_name, levels = display_name[order(Evacuation_Duration_Per_Person)])
  )

# Plot
ggplot(evac_summary, 
       aes(x = Evacuation_Duration_Per_Person, 
           y = Zone_Label, 
           fill = duration_hours)) +
  geom_col() +
  scale_fill_gradient(low = "#EDF3F7", high = "#4A6FA5", name = "Total Duration (hrs)") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Evacuation Duration Per Person by Zone",
    subtitle = "In California",
    x = "Evacuation Duration Per Person (hrs)",
    y = "Evacuation Zone"
  ) +
  theme_minimal(base_size = 13)

```

Next, the median household income for all California block groups was obtained from the U.S. Census site.

B19013_001E was being read as character, but we need it to be numeric for our calculations. Also, there were some NAs, so those were carefully removed. 

Next, we need to link the cleaned ACS income data to the evacuation zones. To do this we will

1. Link ACS median income to each evacuation zone using block group polygons.
2. Aggregate income at the zone level (mean income of affect block groups).
3. Rank zones to see which communities are most economically affected.

```{r check_col_names_evac_zones, eval = TRUE}
colnames(evac_zones_sf)
```


```{r median_household_income_evacuation_zone_integration, eval = TRUE}

library(sf)
library(dplyr)
library(tidycensus)

# Step 1: Load CA block groups
ca_bg <- st_read("tl_2025_06_bg.shp")

# Step 2: Transform to WGS84 (EPSG:4326)
ca_bg <- st_transform(ca_bg, 4326)

# Step 3: Pull median household income from ACS
acs_income <- get_acs(
  geography = "block group",
  variables = "B19013_001",
  state = "CA",
  year = 2021,
  survey = "acs5",
  geometry = FALSE
)

# Step 4: Clean ACS income
acs_income_clean <- acs_income %>%
  select(GEOID, median_income = estimate) %>%
  filter(!is.na(median_income)) %>%
  mutate(median_income = as.numeric(median_income))

# Step 5: Join ACS income to CA block groups
ca_bg$GEOID <- as.character(ca_bg$GEOID)
acs_income_clean$GEOID <- as.character(acs_income_clean$GEOID)

ca_bg_income <- ca_bg %>%
  left_join(acs_income_clean, by = "GEOID")

# Step 6: Transform evacuation zones to match CA block groups
cali_longest_zones_sf <- st_transform(cali_longest_zones_sf, st_crs(ca_bg_income))

```
The shapefile originally used NAD83 (common for TIGER/Line), but the other data (cali_longest_zones_sf) is in WGS84 (EPSG:4326).

st_transform() reprojects the polygons, converting coordinates so everything lines up properly on a map.

We queried the ACS 5-year survey for median household income for all block groups in California. We only kept the GEOID and renamed estimate as median_income. NAs were removed, and median_income was converted to numeric. 

Income data was attached to each blcok group polygon by GEOID, and teh evacuation zones were reprojected to match the CRS of the block groups. 

Next, we will calculate the median household income per evacuation zone and rank the zones based on their income. 

```{r top_economic_disruption_zones, eval = TRUE}

library(dplyr)
library(sf)

# --- Step 1: Ensure CRS matches (just in case) ---
ca_bg_income <- st_transform(ca_bg_income, st_crs(cali_longest_zones_sf))

# --- Step 2: Spatial join block groups to evacuation zones ---
evac_bg_join <- st_join(
  ca_bg_income,
  cali_longest_zones_sf %>% select(display_name, duration_hours),
  join = st_intersects,
  left = FALSE
)

# --- Step 3: Aggregate mean income per evacuation zone ---
zone_income <- evac_bg_join %>%
  group_by(display_name, duration_hours) %>%
  summarise(
    mean_income = mean(median_income, na.rm = TRUE),
    .groups = "drop"
  )

# --- Step 4: Combine with population (from previous overlay) ---
zone_income_pop <- zone_income %>%
  st_drop_geometry() %>%  # drop geometry to join safely
  left_join(evac_population_summary, by = "display_name")  # includes affected_population

# --- Step 5: Compute Evacuation Duration Per Person ---
zone_income_pop <- zone_income_pop %>%
  mutate(
    Evacuation_Per_Person = duration_hours / affected_population
  )

# --- Step 6: Rank-based Economic Disruption Score ---
zone_income_pop <- zone_income_pop %>%
  mutate(
    # Rank zones: lower income → higher rank
    income_rank = rank(mean_income, ties.method = "min"),
    income_rank_rev = max(income_rank) - income_rank + 1,
    
    # Rank zones by evacuation per person: higher → higher rank
    evac_rank = rank(Evacuation_Per_Person, ties.method = "min"),
    
    # Composite rank: sum of ranks
    econ_disruption_rank = income_rank_rev + evac_rank
  ) %>%
  arrange(desc(econ_disruption_rank))

# --- Step 7: View top 10 most economically disrupted zones ---
top_econ_zones <- zone_income_pop %>%
  select(
    Zone = display_name,
    Duration_Hours = duration_hours,
    Population_Affected = affected_population,
    Mean_Income = mean_income,
    Evacuation_Per_Person,
    Income_Rank_Rev = income_rank_rev,
    Evac_Rank = evac_rank,
    Economic_Disruption_Rank = econ_disruption_rank
  )

top_econ_zones
```

For each evacuation zone, we calculated the mean of the median household incomes of all block groups that intersect the zone. This gives us a single representative income value per zone. While each block group’s income is itself a median, the zone-level value is an average of these medians, not the exact median of all households within the zone. We then rank zones by this value, so higher-ranked zones have higher average block-group median incomes. This approach uses the mean of block-group medians as a proxy for the overall economic situation in the zone. It is simpler to compute and avoids needing microdata for every household, but it slightly differs from calculating the true median across the entire zone. Note: ‘mean of medians’ ≈ central tendency of incomes, not exact zone median.

To assess the economic impact of evacuations, we calculated a composite Economic Disruption Rank for each evacuation zone by combining measures of economic vulnerability and evacuation burden. First, each zone was ranked by mean household income, with lower-income zones receiving higher scores to reflect greater economic vulnerability (Income_Rank_Rev). Next, we calculated the evacuation burden per person by dividing total evacuation duration by the affected population, and ranked zones so that higher burdens corresponded to higher ranks (Evac_Rank). Finally, the Economic Disruption Rank was computed as the sum of these two ranks, identifying zones that simultaneously experience longer evacuations per person and greater economic vulnerability. This rank-based approach provides an intuitive measure of which communities are most economically disrupted, while minimizing the influence of extreme values and differing units across the underlying metrics.

```{r econ_disruption_barchart, eval = TRUE}
library(ggplot2)
library(scales)
library(dplyr)

# Arrange zones by rank for plotting
zone_plot_data <- zone_income_pop %>%
  arrange((econ_disruption_rank)) %>%
  mutate(Zone_Label = factor(display_name, levels = display_name))

# Horizontal bar chart
ggplot(zone_plot_data, aes(x = econ_disruption_rank, y = Zone_Label, fill = Evacuation_Per_Person)) +
  geom_col() +
  scale_fill_gradient(low = "#EDF3F7", high = "#4A6FA5", name = "Evacuation per Person") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Economic Disruption Rank by Evacuation Zone",
    subtitle = "Composite of low income and evacuation burden",
    x = "Economic Disruption Rank (higher = more disrupted)",
    y = "Evacuation Zone"
  ) +
  theme_minimal(base_size = 12)
```

```{r econ_disruption_scatterplot, eval = TRUE}

library(ggplot2)
library(ggrepel)
library(scales)

# Identify top 3 zones
top3_zones <- zone_income_pop %>%
  slice_max(econ_disruption_rank, n = 7)

# Scatterplot with labels for top 3 zones
ggplot(zone_income_pop, 
       aes(x = mean_income, 
           y = Evacuation_Per_Person, 
           size = affected_population, 
           color = econ_disruption_rank)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    data = top3_zones,
    aes(label = display_name),
    size = 2.5,
    fontface = "bold",
    nudge_y = 0.5
  ) +
  scale_size(range = c(3, 10), name = "Population Affected") +
  scale_color_viridis_c(option = "plasma", direction = -1, name = "Economic Disruption Rank") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Economic Disruption Across Evacuation Zones",
    x = "Mean Household Income ($)",
    y = "Evacuation Duration per Person (hrs)",
    caption = "Lower income and longer evacuation per person → higher disruption"
  ) +
  theme_minimal(base_size = 13)

```


